---
title: "Class 12: Differential Expression Analysis"
author: "Marcos"
format: gfm
editor_options: 
  chunk_output_type: inline
---

# 2. Import countData and colData

```{r}
counts <- read.csv('airway_scaledcounts.csv', row.names = 1)
metadata <- read.csv('airway_metadata.csv')
```

> **Q1.** How many genes are in this dataset?

In this dataset there are `r nrow(counts)` genes.

> **Q2.**.How many control cell lines do we have?

```{r, include=FALSE}
control_cell_lines <- table(metadata$dex)['control']
```

We have `r control_cell_lines` control cell lines.

# 3. Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
metadata[,"dex"] == "control"
```

```{r}
control <- metadata[metadata[,"dex"]=="control",]

control$id
```

```{r}
control.counts <- counts[,control$id]
control.mean <- rowMeans(control.counts)
```

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]
treated.mean <- rowMeans(counts[,treated$id])
```

```{r}
meancounts <- data.frame(control.mean,
                         treated.mean)
colSums(meancounts)
```

```{r}
plot(meancounts, log='xy')
```

```{r}
log2(40/20)
```

To calculate the log2 of the fold change between treated and control...

```{r}
meancounts$log2fc <- log2(meancounts$treated.mean/
                          meancounts$control.mean)
```

To remove zero values

```{r}
zero.vals <- which(meancounts[,1:2] == 0, arr.ind = TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
```

Overexpressed and underexpressed genes

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
table(up.ind)
```

# 4. DESeq2 analysis

First step, loading the library

```{r}
library(DESeq2)
```

```{r}
citation('DESeq2')
```

Let's generate the specific object that DESeq2 needs:

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design = ~dex )


```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
```

```{r}
summary(res, alpha = 0.05)
```

# 6. Data visualization

```{r}
plot(res$log2FoldChange, -log10(res$padj))
```
